---
title: "Peer Assessment 1 of Reproducible Research"
author: "Marcus Mariano"
date: "friday, february 12, 2016"
output: html_document
keep_md: true
---

## Synopsis



## Loading and preprocessing the data

```{r, echo=FALSE, results='hide'}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
```
Loading data

```{r, echo=TRUE}
data <- read.table(unz("activity.zip", "activity.csv"), sep = ",", 
                  header = TRUE, stringsAsFactors = F)
```

Process and transform the data
```{r, echo=TRUE}
## tidy data
dat <- data
## format the date
dat$date <- as.Date(dat$date, "%Y-%m-%d")
## change to NA the 0 values from steps and interval
dat$steps[dat$steps < 1] <- NA

```

## What is mean total number of steps taken per day?

Total number of steps taken per day

```{r, echo=TRUE}
##  calculate total number of steps taken per day 
total_step_day <- tapply(dat$steps, dat$date, sum, na.rm = TRUE) 
```

Histogram of the total number of steps taken each day

```{r, simple-plot, fig.height = 5, fig.align = 'center'}
hist(total_step_day, breaks = 18, xlab="Total number of steps par day",
     main="Histogram of the total number of steps per day",
      xlim=c(0, 22000))
```

Mean of the total number of steps taken per day

```{r}
total_mean_day <- tapply(dat$steps, dat$date, mean, na.rm = TRUE)
total_mean_day
```

Median of the total number of steps taken per day
```{r}
total_media_day <- tapply(dat$steps, dat$date, median, na.rm = TRUE)
total_media_day
```

## What is the average daily activity pattern?

Series plot of the 5-minute interval and the average number of steps
```{r, fig.height = 5, fig.align = 'center'}
interval_day <- tapply(dat$steps, dat$interval, mean, na.rm = TRUE) 
interlist <- unique(dat$interval)

in_st <- data.frame(interval = interlist, steps = interval_day)

plot(in_st$interval, in_st$steps, type = "l", xlab = "5-minute interval per day", 
     ylab = "Average number of steps", 
     main = "Average daily activity pattern")
```

maximum number of steps on 5-minute interval

```{r}
max(interval_day, na.rm = TRUE)

```

## Imputing missing values

Total number of missing values in the dataset
```{r}
sum(is.na(dat))
```

Filling in all of the missing values in the dataset.
```{r}
dat_new  <- dat ## new data

## the mean for that 5-minute interval
dat_new[is.na(dat_new)] <- mean(interval_day, na.rm = TRUE) ## fill NA's per mean steps per day
```

New dataset with the missing data filled in
```{r}
head(dat_new, 20)
```

Histogram of the total number of steps taken each day
```{r, echo=FALSE, fig.height = 5, fig.align = 'center'}
steps_mean_day <- tapply(dat_new$steps, dat_new$date, sum)

hist(steps_mean_day, breaks = 18, xlab="Total number of steps per day",
     main="Histogram of the total number of steps per day")
##ggplot(steps_mean_day, aes(x = )) + geom_histogram(binwidth = 15)
```

Calculate mean and median total number of steps taken per day
```{r}
new_mean_day <- tapply(dat_new$steps, dat_new$date, mean, na.rm = TRUE)
new_mean_day ## mean total number of steps taken per day

new_media_day <- tapply(dat_new$steps, dat_new$date, median, na.rm = TRUE)
new_media_day ## media total number of steps taken per day
```

Do these values differ from the estimates from the first part of the assignment? 

- Compare to the first part of the assignment the addition of imputing missing data decrease the mean increases the median

Total mean and median from first part of the assignment respectively
```{r, echo=FALSE}
mean(dat$steps, na.rm = TRUE)
median(dat$steps, na.rm = TRUE)
```
Total mean and median from second part of the assignment respectively
```{r, echo=FALSE}
mean(dat_new$steps, na.rm = TRUE)
median(dat_new$steps, na.rm = TRUE)
```

What is the impact of imputing missing data on the estimates of the total daily number of steps?

- As expected the addition of imputing missing data increases the amount of daily steps second part of the assignment, as can be seen below:

Total daily number of steps from first part of the assignment
```{r, echo=TRUE}
##  calculate total number of steps taken per day 
total_step_day <- tapply(dat$steps, dat$date, sum, na.rm = TRUE) 
total_step_day
```

Total daily number of steps from second part of the assignment
```{r, echo=TRUE}
##  calculate total number of steps taken per day 
total_step_day_new <- tapply(dat_new$steps, dat_new$date, sum, na.rm = TRUE) 
total_step_day_new
```

## Are there differences in activity patterns between weekdays and weekends?

Dataset with two levels factor variable - "weekday" and "weekend"
```{r, echo=TRUE}

week <- weekdays(dat$date, abbreviate = TRUE)

translator_function = function(element) {
    switch(element,
           "seg" = "weekday",
           "ter" = "weekday",
           "qua" = "weekday",
           "qui" = "weekday",
           "sex" = "weekday",
           "sáb" = "weekend",
           "dom" = "weekend")
}
week_sapply = sapply(week, translator_function)

week_sapply <- factor(week_sapply, levels = c("weekday", "weekend"))

dat <- dat %>%
    mutate(week = week_sapply) 

head(dat, 12)
```

Panel plot containing a time series plot of averaged across all weekday
```{r, fig.height = 5, fig.align = 'center'}
dat_df <- dat %>%
    filter(week == "weekday") 

interval_day_df <- tapply(dat_df$steps, dat_df$interval, mean, na.rm = TRUE) 
interlist_df <- unique(dat_df$interval)

df <- data.frame(interval = interlist_df, steps = interval_day_df)

plot(df$interval, df$steps, type = "l", xlab = "5-minute interval per day", 
     ylab = "Average number of steps", 
     main = "Average daily activity pattern")
```
